\Large \bf Compilers Project: PTX Super-optimization

### 2018-04-29 paper/main.tex
report writing, slow code-gen

\section{Benchmarks}

@@

@@

\section{Milestones \& Timeline}

In order to progress on this project over the semester,
the following time-line would be paramount:
\begin{itemize}
    \item $20^{th}$ Feb.: Proposal received and reviewing starts.
    \item $27^{th}$ Feb.: Finished installation of nvvm, hiCUDA, and other relevant tools. Download and finding relevant benchmarks.
    \item $6^{th}$ Mar.: Successful encoding of a few bitwise PTX instructions. Testing to be done with samples from hacker's delight.
    \item $13^{th}$ Mar.: A few more instructions.
    \item $20^{th}$ Mar.: Research into conditional verification and implications on PTX.
    \item $27^{th}$ Mar.: Some working conditional test cases.
    \item $3^{rd}$ Apr.: More progress on conditional cases.
    \item $10^{th}$ Apr.: Beginning to encode process larger benchmarks (like those from \cite{hicuda}).
    \item $17^{th}$ Apr.: Debugging, understanding the implications of code size on the optimisation.
    \item $24^{th}$ Apr.: Debugging, understanding the implications of code size on the optimisation. Time permitting, investigation into better heuristics for optimisation.
\end{itemize}

This may be ambitious depending on the complexities
of PTX, so some of the larger benchmarks may not be
fully tested against. The minimal hope is some
optimisation of conditionally evaluated instructions
and straight-line code.

>>>>

The project primarily aims to provide a JIT-compiler for CUDA
so that CUDA can be launched on the fly. However, this goal
leads to utilities for a super-optimiser, especially an object-oriented
IR that can be parsed by any processing layers.

A super-optimizer would be one of these intermediate passes.

\section{The Program}

The program was completed in python for rapid prototyping and testing.
The algorithms, being extensible, should be ported to another language
for speed (CUDA springs to mind due to the data-parallelism in many places)
and tune-ups will be discussed as future work.

There various key modules: a parser, a semantic interpreter, and a code-generator
form the core program.

\subsection{The Parsing Structures and Semantics}

\texttt{pyparsing} is the main library this program rests on.
Assuming just the instructions, the program can read and parse
the most basic arithmetic instructions, handling a few hundred.

Once the basic parsing is completed, the semantics are inferred from the program.
The key structure used is an environment, a 4-tuple:
\begin{itemize}
    \item A map from output to a condition on input.
    \item The set of inputs.
    \item The set of outputs.
    \item The set of all variables ``seen''.
\end{itemize}

Each instruction is also encoded as a parser for its arguments
and a function that takes an environment and updates it to
include the execution of the instruction.

Generally, if the environment read in is $(M, I, O, V)$ and
an instruction $o = f(T)$ (where $T$ is some set of inputs)
is read, the environment is updated as follows:
\begin{itemize}
    \item $o \mapsto f(M(T))$ is added to the map (or $o$ is updated).
        $M(T)$ is used in place of $T$ to propagate equations inferred
        as the program is read.
    \item $(T \setminus (V \cup \{o\})) \cup I$
        This removes all variables assigned to before since they
        cannot provide input to the program. Otherwise, unseen variables
        and past inputs are considered inputs.
    \item $(O \setminus T) \cup \{o\}$
        We remove the inputs to this instruction from the output set of
        the program since we assume that all outputs will not be re-used.
        As a workaround, idempotent instructions may be used to produce the
        output that is re-used. The addition of a simple \texttt{add.u32 actual\_out, temporary\_to\_be\_reused, 0;} would suffice.
    \item $V \cup T \cup \{o\}$.
        We maintain this for the second update above: it is indispensable to
        know that variables have actually been seen in order to determine
        which variables are new inputs.
\end{itemize}

To clarify, here is an example:
\begin{lstlisting}
add.u32 c, a, b;
sub.u32 d, c, b;
add.u32 e, f, g;
\end{lstlisting}
This means that the unsigned integer variables
\texttt{a} through \texttt{f} are computed as follows:
\begin{lstlisting}
c = a + b;
d = c - b;
e = f + g;
\end{lstlisting}
It is clear that \texttt{d = a} and \texttt{c} is not really an
output, more a temporary variable, seen through the lens of the third update
rule above.

This program is parsed into the following 4-tuple:
\begin{align*}
    (e \mapsto f + g, d \mapsto a + b - b), \\
    \{a, b, c, f, g\}, \\
    \{d, e\}, \\
    \{a\ \text{through } f\}
\end{align*}

\subsection{Naive Code-generation}

In order to generate every possible program, a brute-force
enumeration is produced. We assume that each instruction
produces only 1 output (this suffices for all instructions
parsed so far).

The first step of the brute-force approach is to try to satisfy the
spec with an idempotent program that does not alter the variables.
After this, all programs with at least as many instructions as the
outputs provided are run through, adding instructions one at a time.

This approach is incredibly naive: in synthesizing even two instruction
programs, a lot of time is taken as every two-instruction program is
checked without any pruning.

We make this algorithm a little smarter by applying it to only one output
at a time.

\subsection{An Output-by-output Search}

To search the graph of updates, we look at the program output-by-output.
This does not require as many ``restarts'' as a naive traversal
since the instructions and temporary variables synthesized for
one output can be used for another.

We illustrate the algorithm by running through it on the specification from above:
\begin{align*}
    (e \mapsto f + g, d \mapsto a + b - b), \\
    \{a, b, c, f, g\}, \\
    \{d, e\}, \\
    \{a\ \text{through } f\}
\end{align*}

The output-by-output search arbitrarily begins by searching through
all means of generating \texttt{e} and quickly finds \texttt{add.u32 e, f, g;}
finishing its search.

The search for \texttt{add.u32 d, a, 0;} is complicated by potential encoding
troubles that will be mentioned later. Instead of producing the simple
identity, the program complicates matters:
\begin{lstlisting}
sub.u32 t0, f, f;
add.u32 d, a, t0;
\end{lstlisting}
This may be due to semantics in bit-vector theory that require
further investigation.

In order to guarantee that the program produced is in fact the shortest,
the algorithm must be run on all the permutations of the output variables.
Inductively, we can show that the search first generates the shortest
program for an output. Then given the program on $n$ outputs, we search
for the $n + 1^{st}$ output by using the specification of the program of
$n$ outputs. Hence, given an order of outputs, we can be assured that program
produced is the shortest. Hence, a search over all permutations guarantees the
brevity of the program.

\section{Benchmarks and Observations}

@@

This is organized by the general facet of the program that would be improved.
The last subsections include other general comments and ideas, the first subsection
focussing on implementation, the second on theory, and the final on other benchmarks.

\subsection{Parsing}

Currently, most of PTX is not even parsed. This is since straight-line
code is the only facet supported.

A future extension would not only support more instructions but also more
of the structures in PTX. Conditional guards, branches, calls, directives,
and special identifiers are either not supported or ignored.

This limits capabilities in memory optimization. \cite{CUDALoops} suggests
the advantages of also correctly laying memory out.

Furthermore, loops are a key component of CUDA's usage, hence this program
is heavily hindered in its usefulness by not treating loops and conditional
structures. This is discussed more in the next sections.

\subsection{Semantics}

The semantics are lacking in two key ways: type information is not fully
stored for any instruction and the ``environment'' structure discussed
above is inflexible in certain cases.

The lack of typing information makes it difficult to handle instructions such
as \texttt{mul.wide}, where the input and output types vary. Branching instructions
incur such troubles as well as a ``predicate'' type would have to be included.

Furthermore, the environment structure would need to be reasoned about further
in order to handle outputs being re-used. This information may have to be
included in post-conditions.

Additionally, for branching, remembering environments may aid the reading
and understanding of a program.
This may also allow a sort of ``divide-and-conquer'' approach, though programs
so generated are optimal given the division of the input, so all divisions would
have to be considered (much like how all permutations are currently considered).

\subsection{Implementation}

The implementation is in python. Other than documentation and obvious
refactoring, the implementation should be ported for usability.

Python is not only notoriously slow, but is harder to integrate in the
NVCC framework. Hence, compiling with the super-optimiser could be made more
streamlined. A framework such as \cite{GPUOcelot} the Ocelot project
should be considered.

Furthermore, for runtime improvements, other programming languages should be
considered. In fact, the data-parallelism suggests that CUDA should be looked into
as a language to implement the program in. The output-by-output algorithm
could see permutations in parallel and a faster SAT-solver may be usable.

\subsection{Theoretical}

Currently, very few ideas from \cite{superopt} are used since it is
unclear how to apply the graph search. The output-by-output algorithm is
a step towards some of the LENS algorithm, however, backward generation
and the pruning of prior programs are not used. It is unclear how to use
them.

In addition, by finding windows with tighter pre and post-conditions,
the program may hope to find better code segments
that work in the context of the current program.
This would be very useful for CUDA programs given
the conditional guards built into PTX: the preconditions
might be able to be inferred from guards so that
the program produced is better in the context
of the larger code segment.

However, the conditional hazards of CUDA -- that
PTX instructions can be prevented from running
at the instruction level, means that in order to
optimise the bulk of the PTX ISA, conditional
branching cannot be ignored. This would be
a key extension on the original program as
it is a limitation of the current implementation.

@@
